---
output: html_document
editor_options: 
  chunk_output_type: console
---
#------------------------------------------------------------------------------#
# script to test rothc with data from literature research and database: EJPSoil_BUFACAP_template.xlsx
#
# edited by T. Ruhnau (22.10.2024)
#------------------------------------------------------------------------------#

Read packages and input data -> data comes from 02_climate_data.R and 01_auto_database_FB.R
This is the first try to automatize the process of modeling...
```{r Packages}
## load packages
library(SoilR)
library(readr)
library(ggplot2)

# Read dataframes from both scripts before continue (first parts)
# > 02_climate_data_FB.R holds all important data on climate
# > 01_auto_database_FB.R holds all important data on soils
```

```{r Dumpster}
#!# values from ERA5 tiff are all negative, calculation error?
## seems like evaporation = negative values and condensation = positive values according to ERA5 information
## all values are in m > *-1000 to convert to positive values in mm

# Original values from ERA5
# Evp=data.frame(Month=1:12, Evp=c(-0.3, -0.3, -0.3, -0.3, -0.26, -0.24, -0.19, -0.15, -0.15, -0.22, -0.26, -0.32))

# ## plot results
# matplot(years_Spinup, Ct_Spinup, type="l", lty=1, col=1:5,
#         xlab="Time (years)", ylab="C stocks (Mg/ha)")
# legend("topleft", c("DPM", "RPM", "BIO", "HUM", "IOM"),
#        lty=1, col=1:5, bty="n")
```

-- Spin-up section -------------------------------------------------------------

The information of vegetation periods has to be taken directly from the literature or NDVI values, if not available in LTFEs

Split climate data up for historical spin-up and prediction model 
(The split-up assignment 1970-1990 and 1991-2024 is random and follows no peer-reviewed basis)

According to Peña Cabriales et al. (2006) on pg. 59 it says: "one rainy season per year, starting in May-June and ending in October." Filter is set accordingly.
```{r Climate Monthly Means}
# Monthly means from climate_data_FB.R
mm1 <- read.csv2("output/farako_ba/climate/mm_1970_1990.csv", header = TRUE) %>% select(-1)
mm2 <- read.csv2("output/farako_ba/climate/mm_1991_2024.csv", header = TRUE) %>% select(-1)

# Filter Month and Experiment for Section1 Rain
mm1_sec1 <- mm1 %>%
  filter(Month >= 5 & Month <= 10) %>% 
  filter(Experiment == "Farako-Ba") # Choose the experiment

# Filter Month and Experiment for Section2 Dry
mm2_sec2 <- mm2 %>%
  filter(Month %in% c(11:12, 1:4)) %>% # to filter months through year
  filter(Experiment == "Farako-Ba") # Choose the experiment
```

Section1 Rainy Season (May-Oct)
```{r Sec.1 Climate}
# Climate data for Farako-ba, Burkina Faso for 1970-1990 (ERA5)
Temp_Spinup <- mm1_sec1 %>% 
  filter(Variable == "Temp") %>% 
  select(Month, MeanValue) %>% 
  rename(Temp = MeanValue)

Prec_Spinup <- mm1_sec1 %>% 
  filter(Variable == "Prec") %>% 
  select(Month, MeanValue) %>% 
  rename(Prec = MeanValue)

Evap_Spinup <- mm1_sec1 %>% 
  filter(Variable == "Evapo") %>% 
  select(Month, MeanValue) %>% 
  rename(Evap = MeanValue)
```

Section2 Dry (Nov-Apr)
```{r Sec.2 Climate}
## Climate data for Farako-ba, Burkina Faso for 1970-1990 (ERA5)
Temp_Spinup2 <- mm2_sec2 %>% 
  filter(Variable == "Temp") %>% 
  select(Month, MeanValue) %>% 
  rename(Temp = MeanValue)

Prec_Spinup2 <- mm2_sec2 %>% 
  filter(Variable == "Prec") %>% 
  select(Month, MeanValue) %>% 
  rename(Prec = MeanValue)

Evap_Spinup2 <- mm2_sec2 %>% 
  filter(Variable == "Evapo") %>% 
  select(Month, MeanValue) %>% 
  rename(Evap = MeanValue)
```

>> CHECK IF VEGETATION PERIOD IS RIGHT

Section1: Vegetation (Rainy Season) = TRUE (bare = FALSE)
Section2: Vegetation (Dry Season) = FALSE (bare = TRUE)

The idea is, to divide the model calibration into two steps. First step, only for the months with vegetation (Section1) and second step, only for months without vegetation (Section2). 
The respective months should be derived from NDVI or literature values, to make sure during the months there is actually no vegetation covering soils. This also affects Cinputs during the same time (probably around 0). After Section1 & 2 there should be a Section3 combining both models to generate output. 

Section1: Rainy Season
```{r Sec.1 Soils}
# soil data
soil.thick_Spinup = 20   # Soil thickness (organic layer topsoil), in cm
SOC_Spinup = 23.6        # Soil organic carbon in Mg/ha 
clay_Spinup = 13.6       # Percent clay
Cinputs_Spinup = 3       # Annual C inputs to soil in Mg/ha/yr
DPM.RPM = 1.44         # DPM/RPM --> DR = arable: 1.44, grasslands: 0.67, forest: 0.25 (Gottschalk et al. 2012)
FYM = 0                # FYM allocation: DPM: 49%, RPM: 49%, HUM: 2% (Weihermueller et al. 2013)

# create vector of time steps to run simulation
years_Spinup = seq(1/6,500,by=1/6) # change sequence according to rain season
```

Section2: Dry Season (only Cinputs are changing)
```{r Sec.2 Soils}
# soil data
soil.thick_Spinup = 20   # Soil thickness (organic layer topsoil), in cm
SOC_Spinup = 23.6        # Soil organic carbon in Mg/ha 
clay_Spinup = 13.6       # Percent clay
Cinputs_Spinup2 = 1      # Annual C inputs to soil in Mg/ha/yr
DPM.RPM = 1.44         # DPM/RPM --> DR = arable: 1.44, grasslands: 0.67, forest: 0.25 (Gottschalk et al. 2012)
FYM = 0                # FYM allocation: DPM: 49%, RPM: 49%, HUM: 2% (Weihermueller et al. 2013)

# create vector of time steps to run simulation
years_Spinup2 = seq(1/6,500,by=1/6) # change sequence according to vegetation period
```


```{r Spin-Up}
#------------------------------- Section1 -------------------------------------#
## calculate effects of climate on decomposition
fT_Spinup=fT.RothC(Temp_Spinup[,2]) #Temperature effects per month 
fW_Spinup=fW.RothC(P=(Prec_Spinup[,2]), E=(Evap_Spinup[,2]), #!# note: if value is negative, RothCModel cant work
            S.Thick = soil.thick_Spinup, pClay = clay_Spinup,
            pE = 1.0, bare = FALSE)$b #Moisture effects per month

# Replace negative values in fW with 0
fW_Spinup[fW_Spinup < 0] <- 0 # high Evapo values leads to error in model

xi.frame_Spinup=data.frame(years_Spinup,rep(fT_Spinup*fW_Spinup,length.out=length(years_Spinup)))
#--------------------Section2--------------------------------------------------#
### Second spin-up (non-vegetated)

## calculate effects of climate on decomposition
fT_Spinup2=fT.RothC(Temp_Spinup2[,2]) #Temperature effects per month 
fW_Spinup2=fW.RothC(P=(Prec_Spinup2[,2]), E=(Evap_Spinup2[,2]), #!# note: if value is negative, RothCModel cant work
            S.Thick = soil.thick_Spinup, pClay = clay_Spinup,
            pE = 1.0, bare = TRUE)$b #Moisture effects per month

# Replace negative values in fW with 0
fW_Spinup2[fW_Spinup2 < 0] <- 0

xi.frame_Spinup2=data.frame(years_Spinup2,rep(fT_Spinup2*fW_Spinup2,length.out=length(years_Spinup2)))
#--------------------Section_End-----------------------------------------------#

#!# not clear about this method
FallIOM=0.049*SOC_Spinup^(1.139) #IOM using Falloon method

## run the model Section1
## DR = DPM/RPM Ratio must be changed for forest or agroforestry
Model_Spinup=RothCModel(t=years_Spinup,C0=c(DPM=0, RPM=0, BIO=0, HUM=0, IOM=FallIOM), 
                  In=Cinputs_Spinup, clay=clay_Spinup, xi=xi.frame_Spinup,
                  DR = DPM.RPM, FYM = FYM) #Loads the model
Ct_Spinup=getC(Model_Spinup) #Calculates stocks for each pool per month

## run the model Section2
## DR = DPM/RPM Ratio must be changed for forest or agroforestry
Model_Spinup2=RothCModel(t=years_Spinup2,C0=c(DPM=0, RPM=0, BIO=0, HUM=0, IOM=FallIOM), 
                  In=Cinputs_Spinup2, clay=clay_Spinup, xi=xi.frame_Spinup2,
                  DR = DPM.RPM, FYM = FYM) #Loads the model
Ct_Spinup2=getC(Model_Spinup2) #Calculates stocks for each pool per month

## final pool size
poolSize_Spinup=as.numeric(tail(Ct_Spinup,1))
names(poolSize_Spinup)<-c("DPM", "RPM", "BIO", "HUM", "IOM")
poolSize_Spinup

## final pool size Section2
poolSize_Spinup2=as.numeric(tail(Ct_Spinup2,1))
names(poolSize_Spinup2)<-c("DPM", "RPM", "BIO", "HUM", "IOM")
poolSize_Spinup2

# Sum the carbon stocks for test spin up
total_C_spinup <- rowSums(Ct_Spinup)
total_C_spinup2 <- rowSums(Ct_Spinup2)

# Calculate the average total SOC stock over the spin-up period
average_total_SOC_spinup <- mean(total_C_spinup)
average_total_SOC_spinup2 <- mean(total_C_spinup2)

# Display the result
average_total_SOC_spinup
average_total_SOC_spinup2

# Create a data frame for the final pool sizes
final_pools_df <- data.frame(
    Pool = c("DPM", "RPM", "BIO", "HUM", "IOM"),
    Vegetated = poolSize_Spinup,
    NonVegetated = poolSize_Spinup2
)

# Combine total carbon stocks into a data frame
total_C_df <- data.frame(
    Model = c("Vegetated", "Non-Vegetated"),
    Total_C = c(sum(poolSize_Spinup), sum(poolSize_Spinup2))
)

# Determine the maximum y-axis limit to fit both lines comfortably
max_C_spinup <- max(total_C_spinup)
```

Plots results
```{r Plot Spin-Up}
# Plot the final pool sizes
ggplot(final_pools_df, aes(x = Pool)) +
    geom_bar(aes(y = Vegetated, fill = "Vegetated"), stat = "identity", position = "dodge") +
    geom_bar(aes(y = NonVegetated, fill = "Non-Vegetated"), stat = "identity", position = "dodge") +
    labs(title = "Final Pool Sizes for Vegetated and Non-Vegetated Months",
         x = "Carbon Pool",
         y = "Carbon Stock (Mg/ha)",
         fill = "Model") +
    theme_minimal()

# Plot total carbon stocks
ggplot(total_C_df, aes(x = Model, y = Total_C, fill = Model)) +
    geom_bar(stat = "identity") +
    labs(title = "Total Carbon Stocks for Vegetated and Non-Vegetated Months",
         x = "Model",
         y = "Total Carbon Stock (Mg/ha)") +
    theme_minimal()

# Plot individual pools and total SOC for the spin-up period
matplot(years_Spinup, Ct_Spinup, type = "l", lty = 1, col = 1:5,
        xlab = "Time (years)", ylab = "C stocks (Mg/ha)",
        main = "SOC Pools and Total (Spin-Up)", ylim = c(0, max_C_spinup * 1.1))

lines(years_Spinup, total_C_spinup, col = "black", lty = 2)
legend("topright", c("Total SOC", "DPM", "RPM", "BIO", "HUM", "IOM"),
       lty = c(1, 1, 1, 1, 1, 1), col = c(1:5, "black"), bty = "n")
```

Average value for BAU SOC stock = 17.2 t/ha from 1980-2015
Average value for SALM SOC stock = 22.7 t/ha 1980-2015
Therefore average SOC stock = 19.95

-- Prediction section ----------------------------------------------------------
In this section different long-term experiments can be used to predict SOC content in the future.
The data comes from EJPSoil_template database and must be summarized to fit the code.
All climate data comes from ERA5 and must be transformed as mentioned in spin-up run above.
Cinputs must be seperated into business-as-usual (BAU) and project scenario or sustainable agricultural land management (SALM).
With the different inputs, different predictions into future (usually 20 y) will be done.
Clay content and initial SOC must be derived from the respective long-term experiment or strata.
Different long-term experiments are summarized under few strata e.g. IPCC climate zones (CZ) or agro-ecological zones (AEZ).
```{r}
## Climate data for Farako-ba, Burkina Faso for 2023 (ERA5)
Temp=data.frame(Month=1:12, Temp=c(26.2, 28.8, 30.7, 31.2, 30.7, 28.1,
                                   26.5, 25.4, 26.0, 27.4, 28.6, 25.9))
Precip=data.frame(Month=1:12, Precip=c(0.03, 0.06, 6.4, 15.6, 31.9, 110.9, 153.0, 285.3, 104.3,
                                   58.9, 4.9, 0.03))
Evp=data.frame(Month=1:12, Evp=c(350.1, 340.7, 345.9, 333.5, 323.9, 252.8, 203.7, 163.9, 183.5, 216.0,
                                    266.8, 332.8))

## soil data
## taken from Testdata_from_template2.xlsx for 1980 - 2015
soil.thick=20  #Soil thickness (organic layer topsoil), in cm
SOC=19.95       #Soil organic carbon in Mg/ha
clay=13.6        #Same as spin-up, should be edited
Cinputs_BAU=1.2   #Annual C inputs to soil in Mg/ha/yr
Cinputs_SALM=1.9   # Assumed higher C inputs for Project scenario "SALM" = sustainable agriculture land management
DPM.RPM = 1.44      # Standard ratio for arable land, agroforestry and forestry have different standard ratios ???
FYM = 0               # Input factor for Farmyard manure by time

## create vector of time steps to run simulation
years=seq(from=0, to=20, by=1/12) ## changed years from/to > output doesnt change
```

```{r}
## calculate effects of climate on decomposition
fT=fT.RothC(Temp[,2]) #Temperature effects per month
## somehow take care of vegetation period / rain season i.e. bare = FALSe OR TRUE changes during year
fW=fW.RothC(P=(Precip[,2]), E=(Evp[,2]), 
            S.Thick = soil.thick, pClay = clay, 
            pE = 1.0, bare = FALSE)$b #Moisture effects per month

# Replace negative values in fW with 0
fW[fW < 0] <- 0

xi.frame=data.frame(years,rep(fT*fW,length.out=length(years)))

# Falloon is the standard method, other options are Zimmermann et al. (2017) fractionation or pedotransfer functions
FallIOM=0.049*SOC^(1.139) #IOM using Falloon method

## run the model for BAU scenario
Model_BAU=RothCModel(t=years,C0=poolSize_Spinup,
                  In=Cinputs_BAU, clay=clay, xi=xi.frame,
                  DR = DPM.RPM, FYM = FYM) #Loads the model
Ct_BAU=getC(Model_BAU) #Calculates stocks for each pool per month

## run the model for IALM scenario
Model_SALM=RothCModel(t=years,C0=poolSize_Spinup, ## In refers to litter, change FYM if you want to add manure
                  In=Cinputs_SALM, clay=clay, xi=xi.frame) #Loads the model 
Ct_SALM=getC(Model_SALM) #Calculates stocks for each pool per month

# Sum the carbon stocks for each scenario
total_C_BAU <- rowSums(Ct_BAU)
total_C_SALM <- rowSums(Ct_SALM)

# Calculate the difference in total SOC at the end of the time period
difference_total_C <- tail(total_C_SALM, 1) - tail(total_C_BAU, 1)

# Display the result
difference_total_C

# Determine the maximum y-axis limit to fit both lines comfortably
max_C_plot <- max(total_C_SALM, total_C_BAU)
```

Plot results
```{r}
# Plot the total SOC for SALM and BAU scenarios (order changed to SALM first for better line visibility)
plot(years, total_C_SALM, type = "l", col = "red", lty = 2, 
     xlab = "Time (years)", ylab = "Total C stocks (Mg/ha)",
     main = "SOC Stocks Comparison: BAU vs SALM", ylim = c(0, max_C_plot * 1.1))

# Add the BAU line to the plot
lines(years, total_C_BAU, col = "blue", lty = 1)

# Add a legend
legend("topright", legend = c("SALM", "BAU"), col = c("red", "blue"), lty = c(2, 1), bty = "n")

# Display the difference in total SOC at the end on the plot
text(x = max(years) * 0.7, y = 20,
     labels = paste("Delta after 20 years:", round(difference_total_C, 2), "Mg/ha"), pos = 1, col = "black")
```
