#------------------------------------------------------------------------------#
# script to test rothc with data from literature research (EJPSoil_BUFACAP_template.xlsx)
#
# edited by T. Ruhnau (10.07.2024)
#------------------------------------------------------------------------------#

Read packages and input data -> testdata comes from BUFACAP_template
```{r}
## load packages
library(SoilR)

## use condensed dataset from template to run test model (must be modified)
data_test <- readxl::read_excel("C:/Users/t.ruhnau/OneDrive - Unique/Dokumente/SOC-Project/Data/BUFACAP/Testdata_from_template2.xlsx")
data_soil <- readxl::read_excel("C:/Users/t.ruhnau/OneDrive - Unique/Dokumente/SOC-Project/Data/BUFACAP/Test_soil_data.xlsx")
```

-- Spin-up section -------------------------------------------------------------

Climate and soil data for spin-up run, taken from ERA5 and average values from own database (https://uniquelandusede149.sharepoint.com/:x:/s/1_08ProjekteUL2022/ETngkp6ABnpLpmcY818BsJoBIw8DiQxCHEJqeEGT2SelKQ?e=rHU1JQ)
```{r}
## ## Climate data for Farako-ba, Burkina Faso for 1970 (ERA5)
Temp_Spinup=data.frame(Month=1:12, Temp=c(26, 28, 30, 30, 28, 37, 25, 24, 24, 26, 26, 25))
Precip_Spinup=data.frame(Month=1:12, Precip=c(0, 0, 5, 42, 123, 97, 162, 233, 243, 43, 1, 0))

#!# values from ERA5 tiff are all negative, calculation error?
## seems like evaporation = negative values and condensation = positive values according to ERA5 information
## all values are in m > *-1000 to convert to positive values in mm

# Original values from ERA5
# Evp=data.frame(Month=1:12, Evp=c(-0.3, -0.3, -0.3, -0.3, -0.26, -0.24, -0.19, -0.15, -0.15, -0.22, -0.26, -0.32))

# Converted values from ERA5
Evp_Spinup=data.frame(Month=1:12, Evp=c(300, 300, 300, 300, 260, 240, 190, 150, 150, 220, 260, 320))

## soil data
#!# changed values according to average of all experiments in database for 1970-1979
soil.thick_Spinup=20   #Soil thickness (organic layer topsoil), in cm
SOC_Spinup=23.6        #Soil organic carbon in Mg/ha 
clay_Spinup=13.6       #Percent clay
Cinputs_Spinup=3      #Annual C inputs to soil in Mg/ha/yr ##!## estimated value for historical land use from chatgpt
DPM.RPM = 1.44        #DPM/RPM --> DR = arable: 1.44, grasslands: 0.67, forest: 0.25 (Gottschalk et al. 2012)
FYM = 0               # FYM allocation: DPM: 49%, RPM: 49%, HUM: 2% (Weihermueller et al. 2013)

## create vector of time steps to run simulation
years_Spinup=seq(1/12,500,by=1/12) 
```

Spin-up run for test data 
```{r}
## calculate effects of climate on decomposition
fT_Spinup=fT.RothC(Temp_Spinup[,2]) #Temperature effects per month 
fW_Spinup=fW.RothC(P=(Precip_Spinup[,2]), E=(Evp_Spinup[,2]), #!# note: if value is negative, RothCModel cant work
            S.Thick = soil.thick_Spinup, pClay = clay_Spinup,
            pE = 1.0, bare = FALSE)$b #Moisture effects per month

# Replace negative values in fW with 0
fW_Spinup[fW_Spinup < 0] <- 0

xi.frame_Spinup=data.frame(years_Spinup,rep(fT_Spinup*fW_Spinup,length.out=length(years_Spinup)))

#!# not clear about this method
FallIOM=0.049*SOC_Spinup^(1.139) #IOM using Falloon method

## run the model
## DR = DPM/RPM Ratio must be changed for forest or agroforestry
Model_Spinup=RothCModel(t=years_Spinup,C0=c(DPM=0, RPM=0, BIO=0, HUM=0, IOM=FallIOM), 
                  In=Cinputs_Spinup, clay=clay_Spinup, xi=xi.frame_Spinup,
                  DR = DPM.RPM, FYM = FYM) #Loads the model
Ct_Spinup=getC(Model_Spinup) #Calculates stocks for each pool per month

# ## plot results
# matplot(years_Spinup, Ct_Spinup, type="l", lty=1, col=1:5,
#         xlab="Time (years)", ylab="C stocks (Mg/ha)")
# legend("topleft", c("DPM", "RPM", "BIO", "HUM", "IOM"),
#        lty=1, col=1:5, bty="n")

## final pool size
poolSize_Spinup=as.numeric(tail(Ct_Spinup,1))
names(poolSize_Spinup)<-c("DPM", "RPM", "BIO", "HUM", "IOM")
poolSize_Spinup

# Sum the carbon stocks for test spin up
total_C_spinup <- rowSums(Ct_Spinup)

# Calculate the average total SOC stock over the spin-up period
average_total_SOC_spinup <- mean(total_C_spinup)

# Display the result
average_total_SOC_spinup

# Determine the maximum y-axis limit to fit both lines comfortably
max_C_spinup <- max(total_C_spinup)
```

Plots results
```{r}
# Plot individual pools and total SOC for the spin-up period
matplot(years_Spinup, Ct_Spinup, type = "l", lty = 1, col = 1:5,
        xlab = "Time (years)", ylab = "C stocks (Mg/ha)",
        main = "SOC Pools and Total (Spin-Up)", ylim = c(0, max_C_spinup * 1.1))

lines(years_Spinup, total_C_spinup, col = "black", lty = 2)
legend("topright", c("Total SOC", "DPM", "RPM", "BIO", "HUM", "IOM"),
       lty = c(1, 1, 1, 1, 1, 1), col = c(1:5, "black"), bty = "n")
```

Average value for BAU SOC stock = 17.2 t/ha from 1980-2015
Average value for SALM SOC stock = 22.7 t/ha 1980-2015
Therefore average SOC stock = 19.95

-- Prediction section ----------------------------------------------------------
In this section different long-term experiments can be used to predict SOC content in the future.
The data comes from EJPSoil_template database and must be summarized to fit the code.
All climate data comes from ERA5 and must be transformed as mentioned in spin-up run above.
Cinputs must be seperated into business-as-usual (BAU) and project scenario or sustainable agricultural land management (SALM).
With the different inputs, different predictions into future (usually 20 y) will be done.
Clay content and initial SOC must be derived from the respective long-term experiment or strata.
Different long-term experiments are summarized under few strata e.g. IPCC climate zones (CZ) or agro-ecological zones (AEZ).
```{r}
## Climate data for Farako-ba, Burkina Faso for 2023 (ERA5)
Temp=data.frame(Month=1:12, Temp=c(26.2, 28.8, 30.7, 31.2, 30.7, 28.1,
                                   26.5, 25.4, 26.0, 27.4, 28.6, 25.9))
Precip=data.frame(Month=1:12, Precip=c(0.03, 0.06, 6.4, 15.6, 31.9, 110.9, 153.0, 285.3, 104.3,
                                   58.9, 4.9, 0.03))
Evp=data.frame(Month=1:12, Evp=c(350.1, 340.7, 345.9, 333.5, 323.9, 252.8, 203.7, 163.9, 183.5, 216.0,
                                    266.8, 332.8))

## soil data
## taken from Testdata_from_template2.xlsx for 1980 - 2015
soil.thick=20  #Soil thickness (organic layer topsoil), in cm
SOC=19.95       #Soil organic carbon in Mg/ha
clay=13.6        #Same as spin-up, should be edited
Cinputs_BAU=1.2   #Annual C inputs to soil in Mg/ha/yr
Cinputs_SALM=1.9   # Assumed higher C inputs for Project scenario "SALM" = sustainable agriculture land management
DPM.RPM = 1.44      # Standard ratio for arable land, agroforestry and forestry have different standard ratios ???
FYM = 0               # Input factor for Farmyard manure by time

## create vector of time steps to run simulation
years=seq(from=0, to=20, by=1/12) ## changed years from/to > output doesnt change
```

```{r}
## calculate effects of climate on decomposition
fT=fT.RothC(Temp[,2]) #Temperature effects per month
## somehow take care of vegetation period / rain season i.e. bare = FALSe OR TRUE changes during year
fW=fW.RothC(P=(Precip[,2]), E=(Evp[,2]), 
            S.Thick = soil.thick, pClay = clay, 
            pE = 1.0, bare = FALSE)$b #Moisture effects per month

# Replace negative values in fW with 0
fW[fW < 0] <- 0

xi.frame=data.frame(years,rep(fT*fW,length.out=length(years)))

# Falloon is the standard method, other options are Zimmermann et al. (2017) fractionation or pedotransfer functions
FallIOM=0.049*SOC^(1.139) #IOM using Falloon method

## run the model for BAU scenario
Model_BAU=RothCModel(t=years,C0=poolSize_Spinup,
                  In=Cinputs_BAU, clay=clay, xi=xi.frame,
                  DR = DPM.RPM, FYM = FYM) #Loads the model
Ct_BAU=getC(Model_BAU) #Calculates stocks for each pool per month

## run the model for IALM scenario
Model_SALM=RothCModel(t=years,C0=poolSize_Spinup, ## In refers to litter, change FYM if you want to add manure
                  In=Cinputs_SALM, clay=clay, xi=xi.frame) #Loads the model 
Ct_SALM=getC(Model_SALM) #Calculates stocks for each pool per month

# Sum the carbon stocks for each scenario
total_C_BAU <- rowSums(Ct_BAU)
total_C_SALM <- rowSums(Ct_SALM)

# Calculate the difference in total SOC at the end of the time period
difference_total_C <- tail(total_C_SALM, 1) - tail(total_C_BAU, 1)

# Display the result
difference_total_C

# Determine the maximum y-axis limit to fit both lines comfortably
max_C_plot <- max(total_C_SALM, total_C_BAU)
```

Plot results
```{r}
# Plot the total SOC for SALM and BAU scenarios (order changed to SALM first for better line visibility)
plot(years, total_C_SALM, type = "l", col = "red", lty = 2, 
     xlab = "Time (years)", ylab = "Total C stocks (Mg/ha)",
     main = "SOC Stocks Comparison: BAU vs SALM", ylim = c(0, max_C_plot * 1.1))

# Add the BAU line to the plot
lines(years, total_C_BAU, col = "blue", lty = 1)

# Add a legend
legend("topright", legend = c("SALM", "BAU"), col = c("red", "blue"), lty = c(2, 1), bty = "n")

# Display the difference in total SOC at the end on the plot
text(x = max(years) * 0.7, y = 20,
     labels = paste("Delta after 20 years:", round(difference_total_C, 2), "Mg/ha"), pos = 1, col = "black")
```
